# Coursera Reproducible Research Course

# Assignment 1

In this assignment we will analyse the number of steps recorded in five
minute intervals during the period of 61 days from Monday 1st Oct 2012 to 
Friday 30th Nov 2012 by people wearing activity monitoring devices.

The dataset used is the Activity Monitoring Data : 'activity.csv' 

## Section A : Loading and Preprocessing the data

The data archive was downloaded from the course website and the file 
**activity.csv** unzipped into the project folder. 

The R code to read the file is :-


```{r}
Data <- read.csv("activity.csv")
str(Data)

```
Examining the downloaded file (using the function *View(Data)*) 
indicates that the file has a regular structure with counts of steps 
for 288 five minute intervals in chronological order for each of the 
61 days. (61 x 288 = 17,568 rows of data).

There are three variables :

- Numbers of steps in each interval (steps : integer)
- Date (date : character string)
- Five minute interval identifier : (interval : integer of the form *hhmm*)

There are NA values in some of the 'steps' variables. A visual inspection of 
the data suggests that no counts were taken on the following days :

- Monday    2012/10/01
- Monday    2012/10/08
- Thursday  2012/11/01
- Sunday    2012/11/04
- Friday    2012/11/09
- Saturday  2012/11/10
- Wednesday 2012/11/14
- Friday    2012/11/30

In order to facilitate the analysis, some preprocessing was undertaken :

- An index variable in the range 1:288 was substituted for interval identifiers
- The weekday name and a weekday index number was added to each record
- A new data.frame was created by subsetting to remove any incomplete records (NA's)
- A day number index variable was added to the original and subset data.frames.

The subsetting operation removed all the rows relating to the eight days previously
identified, (8 x 288 = 2304 rows).

**Convert interval variableto index, add day names and index number & subset :**

```{r}
Data$interval <- as.integer(as.factor(Data$interval))
Data$Weekday <- as.factor(weekdays(as.Date(Data$date)))
Data$Weekdaynumber <- as.integer(Data$Weekday)

Temp <- complete.cases(Data)
Subset <- Data[Temp, ]

```
**Add date indexes & determine number of rows and number of complete days removed by subsetting**

```{r}
Data$daynumber <- as.integer(as.factor(as.Date(Data$date)))
Subset$daynumber <- as.integer(as.factor(as.Date(Subset$date))) 
head(Data)
head(Subset)
RR <- dim(Data)[1] - dim(Subset)[1]
DR <- max(Data$daynumber) - max(Subset$daynumber)
Stats <- data.frame(Legend...=c("Rows removed  :", "Days removed  :"), 
		Statistics=c(RR, DR))
Stats
View(Data)
View(Subset)

```
## Section B : Calculate and plot daily total number of steps recorded

This analysis was performed on the subset data with the NA values removed.

### 1. Calculate total number of steps for each of the 53 days in the subset : 

```{r}
Dailysteps <- rep(0, max(Subset$daynumber))
for(i in 1:dim(Subset)[1])
{	Dailysteps[Subset$daynumber[i]] <- Dailysteps[Subset$daynumber[i]] + Subset$steps[i]
}

## For comparison, reinsert the NA values
OriginalDailysteps <- c(NA, Dailysteps[1:6], NA, Dailysteps[7:29], NA, 
		Dailysteps[30:31], NA, Dailysteps[32:35], NA, NA, Dailysteps[36:38],
		NA, Dailysteps[39:53], NA)

```
### 2. Plot the histogram :

```{r}
library(ggplot2)
qplot(Dailysteps, geom="histogram", xlab="Number of steps",
	main="Total number of steps per day", binwidth=500, asp=1.0)

```
### 3. Mean and median of the total number of steps per day :

```{r}
Stats <- data.frame(Mean=round(mean(Dailysteps), 0), Median=median(Dailysteps))
Stats
 
```
## Section C : Investigate mean number of steps per five minute interval

### 1. Calculate mean number of steps in each five minute interval using subset data

```{r}
SubIntervalsteps <- rep(0, max(Subset$interval))
for(i in 1:dim(Subset)[1])
{	SubIntervalsteps[Subset$interval[i]] <- SubIntervalsteps[Subset$interval[i]] + Subset$steps[i]
}
SubIntervalsteps <- SubIntervalsteps / max(Subset$daynumber)
head(SubIntervalsteps)

```
### 2. Create time series plot

The vector of means was converted to an hourly time series with frequency = 12
using function *ts()* and plotted with the generic *plot()* function.

```{r}
SubsetCount <- ts(SubIntervalsteps, start=1, frequency=12)
plot(SubsetCount, main="Daily Mean Number of Steps per Five Minute Interval", 
	xlab="Time of day (hours)")

```
### 3. Find time interval with maximum activity level

```{r}
Max <- max(SubIntervalsteps)
MaxInt <- 0
for(i in 1:288)

{	if(SubIntervalsteps[i] == Max) MaxInt <- i
}
Stats <- data.frame(Statistic=c("Maximum steps", "Interval number"), 
					Value=c(round(Max, 0), MaxInt))
Stats

```
Interval 104 (8.40 am) had greatest activity with a mean of 206 steps.


## Section D : Impute missing values and compare with original analysis

The weekdays with missing data are : Monday (2), Wednesday (1), Thursday (1),
Friday (2), Saturday (1). 

The data collection started on a Monday and included 9 sets of weekdays and 
eight weekends; therefore data was collected for : 
Mondays (7), Tuesdays (9), Wednesdays (8), Thursdays (8), Fridays (7), Saturdays (7), Sundays (8).

To minimise bias we will calculate the mean number of steps per time interval
for each day of the week from the subset data and replace the NA values with 
the mean values for the relevant day.

### 1. Create matrix of mean number of steps in each five minute interval per weekday
 
```{r}
Temp <- matrix(0, 288, 7)
for(i in 1:dim(Subset)[1])
{	Temp[Subset$interval[i], Subset$Weekdaynumber[i]] <- Temp[Subset$interval[i], Subset$Weekdaynumber[i]] + Subset$steps[i]
}
Daycount <- c(7, 7, 7, 8, 8, 9, 8) ## days with measurements
Dayorder <- c(2, 6, 7, 5, 6, 3, 4)	## days are indexed in alphabetical order
for(i in 1:7)
{	Temp[ , i] <- round(Temp[ , i] / Daycount[i], 0)
}

## display a sample

Stats <- data.frame(Mon=Temp[100:105, Dayorder[1]],
			Tue=Temp[100:105, Dayorder[2]],
			Wed=Temp[100:105, Dayorder[3]],
			Thu=Temp[100:105, Dayorder[4]],
			Fri=Temp[100:105, Dayorder[5]],
			Sat=Temp[100:105, Dayorder[6]],
			Sun=Temp[100:105, Dayorder[7]])

Stats

```
### 2. Substitute calculated means for NA's in original data

```{r}
for(i in 1:dim(Data)[1])
{	if(is.na(Data$steps[i])) Data$steps[i] <- Temp[Data$interval[i] , Data$Weekdaynumber[i]]
}
View(Data)

```

### 3. Calculate total number of steps in each five minute interval

```{r}
ImputedDailysteps <- rep(0, max(Data$daynumber))
for(i in 1:dim(Data)[1])
{	ImputedDailysteps[Data$daynumber[i]] <- ImputedDailysteps[Data$daynumber[i]] + Data$steps[i]
}

```
**Total daily number of steps before and after imputing :**

```{r}
OriginalDailysteps
ImputedDailysteps

```

### 4. Construct histograms of the original and imputed data.


```{r}
par(mfrow=c(1, 2))
hist(Dailysteps,  xlab="Number of steps",
	main="Total number of steps per day\n(original data)", breaks=30, col=2)
hist(ImputedDailysteps,  xlab="Number of steps",
	main="Total number of steps per day\n(Imputed data)", breaks=30, col=3)

```

### 5. Mean and median of the total number of steps per day :

```{r}
Origmean <- round(mean(Dailysteps), 0)
Impmean <- round(mean(ImputedDailysteps), 0)
Origmedian <- median(Dailysteps)
Impmedian <- median(ImpDailysteps)

Stats <- data.frame(Statistic=c("Mean    :", "Median  :"),
				Original_data=c(Origmean, Origmedian),
				Imputed_data=c(Impmean, Impmedian))
Stats

```

## Section E : Analysis of weekday / weekend difference in activity patterns

We will calculate and compare the mean number of steps in each five minute 
interval for weekdays and weekends, using the full dataset with imputed values,
and create a multivariate time series plot.

As noted above, the two month period covered by the data covers nine sets of
weekdays (9 x 5 = 45 days) and eight weekends (8 x 2 = 16 days).

**Calculate mean number of steps for weekdays and weekends**
 
```{r}
Weekdaysteps <- Weekendsteps <- rep(0, max(Data$interval))
for(i in 1:dim(Data)[1])
{	if(Data$Weekdaynumber[i] == 3 || Data$Weekdaynumber[1] == 4)
	{ Weekendsteps[Data$interval[i]] <- Weekendsteps[Data$interval[i]] + Data$steps[i] }
	else
	{ Weekdaysteps[Data$interval[i]] <- Weekdaysteps[Data$interval[i]] + Data$steps[i] }
}
Weekdaysteps <- Weekdaysteps / 45
Weekendsteps <- Weekendsteps / 16

```

**Convert to time series and plot**

```{r}
Counts2 <- ts(data.frame(Weekdays=Weekdaysteps, Weekends=Weekendsteps), start=1, frequency=12)
View(Counts2)
plot(Counts2, plot.type="m", mar=c(gap=0.5, 4.1, gap=0.5, 2.1), ylim=c(0, 250),
	main="Mean Number of Steps per Five Minute Interval\nComparison between weekdays & weekends", 
	xlab="Time of day (hours)", col=4)

```

The graph shows that there is generally less activity at the weekends, 
in particular, the early morning peak in activity is less pronounced and 
daytime activity seems to start and end about two hours later in the day at 
weekends. (Note different y-axis scales).


**---------------------------------------------------------------------------------------------------**


```{r, echo=FALSE}
time <- format(Sys.time(), "%a %b %d %Y - %X")

```

#### Created from 'CRR_Project_1.Rmd' on `r time`

